---
title: "largeVis"
output: github_document
---

This is an implementation of the `largeVis` algorithm described in (https://arxiv.org/abs/1602.00370).  It also incorporates code for a very fast algorithm for estimating k-nearest neighbors.

The inner loops for nearest-neighbor search and gradient descent are implemented in C++ using `Rcpp` and `RcppArmadillo`.  (If you get an error that a `NULL value passed as symbol address` this relates to `Rcpp` and please open an issue here.)

#### Project Status & Caveats
* This project is under heavy development.  It is likely that there are still bugs in the math. 
* Currently, the code is tested to perform correctly with small datasets. 
* Performance in the neighborhood exploration phase is unusually slow; I am working on tracing the cause. 

#### Examples:

```{r iris,eval=T}
library(largeVis,quietly=T)
data(iris)
dat <- as.matrix(iris[,1:4])
dat <- scale(dat)
dupes = which(duplicated(dat))
dat <- dat[-dupes,]
visObject <- largeVis(dat, pca.first = F, 
                   max.iter = 20, sgd.batches = 800000, 
                   K = 10,  gamma = 2, rho = 1, M = 40, alpha = 20,verbose=F)
```
```{r echo=F}
coords <- data.frame(visObject$coords)
colnames(coords) <- c("X", "Y")
coords$Species <- iris$Species[-dupes]
coords$algo <- "largeVis"
```
```{r iristsne,echo=F}
library(Rtsne,quietly=T)
more <- Rtsne(dat, initial_dims = 4, pca = F)$Y
more <- data.frame(X = more[,1], Y = more[,2], algo = "B-H t-SNE")
more$Species <- iris$Species[-dupes]
coords <- rbind(coords, more)
coords$algo <- factor(coords$algo)
require(ggplot2)
```
```{r showiris,echo=F}
ggplot(coords, aes(x = X, y = Y, color = Species)) + 
  geom_point(size = 0.5) + 
  facet_grid(~ algo, scales = 'free') + 
  ggtitle("The Iris Dataset largeVis vs. t-SNE")
```


```{r mnist,eval=F,echo=F}
load("../../kaggle/mnist/mnist.Rda")
dat <- mnist$images
dim(dat) <- c(42000, 28 * 28)
dat <- (dat / 255) - 0.5
coords <- largeVis(dat, pca.f = F, n.tree = 10, tree.th = 100, K = 40, sgd = 1e7, alpha = 2, rho = 1)$coords
coords <- data.frame(X = coords[,1], Y = coords[,2], label = mnist$labels)
ggplot(coords, aes(x = X, y = Y, color = label)) + geom_point(size = 0.2, alpha = 0.3) + ggtitle("MNIST")
```
```{r mnist_img,eval=F,echo=F}
require(raster)
samples <- sample(nrow(coords), 5000, replace = F)
toplot <- coords[samples,]
plot(NA, xlim = c(min(toplot[,1]), max(toplot[,1])), ylim = c(min(toplot[,2]), max(toplot[,2])))

for (i in samples) {
  rasterImage(mnist$images[i,,], toplot[i,1],toplot[i,2], toplot[i,1] + 0.02,toplot[i,2] + 0.02 )
}
```
