---
title: "largeVis"
output: github_document
bibliography: vignettes/TangLZM16.bib
---
```{r getversion,eval=T,echo=F,warning=F,error=F,message=F}
branch <- system("git branch -v", intern=TRUE)
branch <- branch[grep("\\*", branch)]
poses <- regexpr("(?<=\\*\\s)(\\S+)(?=\\s)", perl = TRUE, branch)
branch <- substr(branch, attr(poses, "capture.start"), attr(poses, "capture.start") + attr(poses, "capture.length"))
branch <- gsub("^\\s+|\\s+$", "", branch)
```
[![Travis-CI Build Status](https://travis-ci.org/elbamos/largeVis.svg?branch=`r branch`)](https://travis-ci.org/elbamos/largeVis)
[![Coverage Status](https://img.shields.io/codecov/c/github/elbamos/largeVis/`r branch`.svg)](https://codecov.io/gh/elbamos/largeVis/branch/`r branch`) [![https://gitter.im/elbamos/largeVis](https://badges.gitter.im/elbamos/largeVis.svg)](https://gitter.im/elbamos/largeVis?utm_source=badge&utm_medium=badge&utm_campaign=pr-badge&utm_content=badge)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/elbamos/largeVis?branch=`r branch`&svg=true)](https://ci.appveyor.com/project/elbamos/largeVis?branch=`r branch`)

This is an implementation of the `largeVis` algorithm described in (https://arxiv.org/abs/1602.00370).  It also incorporates a very fast algorithm for estimating k-nearest neighbors, implemented in C++ with `Rcpp` and `OpenMP`, and for visualizing a map of the manifold like [this](http://cs.stanford.edu/people/karpathy/cnnembed/).


#### Project Status & Caveats
* In final testing before submission to CRAN.
* Tested with (dense) matrices > 2.5 Million rows, and sparse matrices with > 10,000 features. 
* Performance and memory efficiency are good. 
* I have been able to replicate, in the sense of producing characteristically similar visualizations, the results in the original paper. 

### Notes
* In some configurations on Linux, the final visualizations tend to be "fuzzy." Please contact me if you experience this, I'd like to try to pin-down the source.
* On Mac OS X, the Apple compiler does not support OpenMP.  To compmile `largeVis` with openmp support, use `homebrew` to install `llvm` version 3.8 or greater, then link it to `/usr/local` with `brew link --force llvm`. This may cause issues if you use the Apple compiler for other development, however in that case you should know your configuration well enough to resolve the issue.

## Examples
```{r echo=F}
vdatapath <- "../largeVisData/vignettedata/"
```
```{r child='Examples.Rmd',}
```

## Benchmarks

```{r child='vignettes/benchmarks.Rmd',}
```

## Vignette

```{r echo=F}
vdatapath <- "../../largeVisData/vignettedata/"
```
```{r child='vignettes/largeVis.Rmd',}
```
