---
title: 'largeVis: An Implementation of the LargeVis Algorithm'
author: "Amos Elberg"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    fig_caption: yes
bibliography: TangLZM16.bib
vignette: >
  %\VignetteIndexEntry{largeVis}   
  %\VignetteEngine{knitr::rmarkdown}   
  %\VignetteEncoding{UTF-8}
---

```{r setup,eval=T,echo=F,warning=F,error=F,message=F}
# Note to reader:  Please don't steal the semi-distinctive visual style I spent several minutes creating for myself.
require(ggplot2, 
        quietly = TRUE)
require(RColorBrewer, 
        quietly = TRUE)
require(wesanderson, 
        quietly = TRUE)
library(jpeg,
        quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>")
colors_discrete <- function(x) rep(wes_palette("Darjeeling", 
                                               n = min(x, 5)), 
                                   2)[1:x]
colors_divergent_discrete <- function(x) 
  grDevices::colorRampPalette(RColorBrewer::brewer.pal(x, "Spectral"))
colors_continuous <-  function(x) wes_palette(name = "Zissou",
                                              n = x, 
                                              type = "continuous")

nacol <- colors_discrete(4)[4]
theme_set(
  theme_bw() %+replace%
  theme(
    legend.key.size = unit(4, "mm"), 
    legend.title = element_text(size = rel(0.8),
                              face = "bold"),
    legend.margin = unit(0, "cm"),
    legend.position = "bottom",
    legend.key.size = unit(0.5, "lines"),
    legend.text=element_text(size = unit(8, "points")), 
    axis.title.y = element_text(angle = 90),
    axis.text = element_text(size = rel(0.7)),
    plot.margin = unit(c(0, 0.5, 1, 0), "lines"), 
    axis.title = element_text(size = rel(0.8),
                              face = "bold"),
    title = element_text(size = rel(0.9))
  ) 
)
rebuild <- FALSE
if (!exists("buildManifolds")) buildManifolds <- rebuild

require(largeVis,quietly = TRUE)
```
This Vingette provides an overview of the largeVis package.  

## Introduction

This package provides `LargeVis` visualizations and fast nearest-neighbor search.  The `LargeVis` algorithm, presented in @tang2016visualizing, creates high-quality low-dimensional representaitons of large, high-dimensional datasets, similar to [t-SNE](https://lvdmaaten.github.io/tsne/).  

These visualizations are useful for data exploration, for visualizing complex non-linear functions, and especially for visualizing embeddings such as learned vectors for images. 

A limitation of t-SNE is that because the algorithm has complexity order $O(n^2)$, it is not feasible for use on even moderately sized datasets.  [Barnes-Hut](https://arxiv.org/pdf/1301.3342.pdf), an approximation of t-SNE, has complexity $O(n \log n)$ but also quickly becomes infeasible as the size of data grows. `LargeVis` is intended to address the issue by operating in linear $O(n)$ time.  It has been benchmarked at more than 30x faster than Barnes-Hut on datasets of approximately 1-million rows, and scaled linearly as long as there is sufficient RAM. 

In addition, `LargeVis` includes an algorithm for finding approximate k-Nearest Neighbors in $O(n)$ time. This algorithm turns out to be faster at finding accurate a-NNs than any other method I was able to test. 

The package also includes a function for visualizing image embeddings by plotting images at the locations given by the `LargeVis` algorithm.

For a detailed description of the algorithm, please see the original paper, @tang2016visualizing.

## Package Overview

The `largeVis` package offers four functions for visualizing high-dimensional datasets and finding approximate nearest neighbors (along with some helper functions):

1.  `randomProjectionTreeSearch`, a method for finding approximate nearest neighbors.
2.  `projectKNNs`, which takes as input a weighted nearest-neighbor graph and estimates a projection into a low-dimensional space.
3.  `vis`, which implements the entire `LargeVis` algorithm.
4.  `manifoldMap` (and companon `ggManifoldMap`), which produce a plot for visualizing embeddings of images. 

See the [original paper](https://arxiv.org/abs/1602.00370) for a detailed description of the algorithm. 

## Data Preparation

For input to `largeVis`, data should be scaled, NA's, Infs and NULL removed, and transposed from the R-standard so that examples are columns and features are rows. Duplicates should be removed as well.

If there are NA's, Infs, or NULLs in the input, `randomProjectionTreeSearch` will definitely fail. 

If the numerical range covered by the data is large, this can cause errors in or before the `buildEdgeMatrix` function. This is because the algorithm requires calculating $\exp(||\vec{x_i}, \vec{x_j}||^2)$ in the high-dimensional space, which will overflow if the distance between any nearest neighbors exceeds about 26.  

If there are duplicates in the input data, while the implementation tries to filter duplicates, it is likely to lead to problems. If the number of duplicates is large, this can cause the random projection tree search to fail. If the number is small, the algorithm may identify a sufficient number of neighbors, but an error may then occur during `buildEdgeMatrix`, or stochastic gradient descent. 

## Examples

```{r MNIST,echo=F,message=F,warning=F,results='hide',eval=rebuild}
darch::provideMNIST(download=T)
load("data/train.RData")

mnistCoords <- vis(t(trainData) - 0.5, 
                   K = 40, 
                   tree_threshold = 700, 
                   n_trees = 40, 
                   max_iter = 2, 
                   verbose=F)
mnistCoords <- mnistCoords$coords
mnistCoords <- scale(t(mnistCoords))
mnistCoords <- data.frame(mnistCoords)
colnames(mnistCoords) <- c("x", "y")
labs <- apply(trainLabels, 
              MARGIN = 1, 
              FUN = function(x) which(x == 1))
mnistCoords$labels <- factor(labs - 1)
```

```{r drawmnist,echo=F,warning=F}
load(system.file("extdata", "mnistcoords.Rda", package="largeVis"))
ggplot(mnistCoords, aes(x = x, y = y, color = labels)) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_x_continuous(name = "", limits = c(-2, 2), breaks = NULL) +
  scale_y_continuous(name = "", limits = c(-2, 2), breaks = NULL) +
  scale_color_manual(values = colors_divergent_discrete(10)(10)) +
  guides(colour = guide_legend(override.aes = list(size=5))) +
  ggtitle("MNIST")
```

```{r ldafromldavis,echo=F,eval=rebuild}
library(LDAvis)
data("TwentyNewsgroups")
theta <- TwentyNewsgroups$theta
dd <- duplicated(theta) # This step is very important
theta <- theta[! dd, ]
set.seed(1974)
initialcoords <- matrix(rnorm(2 * ncol(theta)), ncol = ncol(theta))
visObj <- vis(t(scale(theta)), 
              K = 50, 
              n_trees = 20, 
              tree_threshold = 100, 
              max_iter = 1, 
              coords = initialcoords)
library(Rtsne)
tsneCoords <- data.frame(Rtsne(t(theta), pca = FALSE)$Y)
tsneCoords$model <- 'tsne'
ngcoords <- data.frame(scale(t(visObj$coords)))
ngCoords$model <- 'largeVis'
library(lda)
data("newsgroup.train.labels")
ngcoords$label <- factor(newsgroup.train.labels)[-1][! dd]
tsneCoords$label <- factor(newsgroup.train.labels)[-1][! dd]
ngcoords <- rbind(ngcoords, tsneCoords)
colnames(ngcoords) <- c("x", "y", "model")
ngcoords$model <- factor(ngcoords$model)
```
```{r draw20ng,echo=F,warning=FALSE,error=FALSE,message=FALSE}
load(system.file("extdata", "ngcoords.Rda", package = "largeVis"))
ggplot(ngcoords, 
       aes(x = x, 
           y = y, 
           color = label)) +
  geom_point(size = 0.4, alpha = 0.5) + 
  scale_color_discrete(guide = FALSE) +
  scale_x_continuous(name = "", limits = c(-1.7, 2), breaks = NULL) +
  scale_y_continuous(name = "", limits = c(-2, 2.5), breaks = NULL) +
  ggtitle(expression(
    atop("20 Newsgroups LDA Vectors",
         atop(italic("(n = 11268, K = 100, n_trees = 20, threshold = 100)"))
         )
    ))  
```

```{r wikiterms,eval=F,echo=F}
# # The data file must be obtained directly from the paper authors
wikiwords <- readr::read_delim("/mnt/hfsshare/DATASETS/Wiki_embedding/word_vec",delim= " ", col_names = FALSE, skip = 1)
wikiwords <- t(as.matrix(wikiwords[, 2:101]))
set.seed(1974)
initcoords <- matrix(rnorm(ncol(wikiwords) * 2), nrow = 2)
wikiVis <- vis(wikiwords, 
               coords = initcoords, 
               K = 100, 
               tree_threshold = 100, 
               n_trees = 50)
wikiwords <- readr::read_delim("/mnt/hfsshare/DATASETS/Wiki_embedding/word_vec",delim= " ", col_names = FALSE, skip = 1)
wikiwords <- wikiwords[,1]$X1
wikilabels <- readr::read_delim("/mnt/hfsshare/DATASETS/Wiki_embedding/wiki_word_label.txt", 
                                col_names = FALSE, delim = "\t")
words <- data.frame(word = wikiwords, 
                    labelrow = match(wikiwords, wikilabels$X1))
words$label <- wikilabels$X2[words$labelrow]
words <- cbind(words, 
               data.frame(t(wikiVis$coords)))
colnames(words) <- c('word', 'labelrow', 'label', 'x', 'y')
```

```{r drawwikiwords,echo=F,eval=T,warning=F,message=F}
load(system.file("extdata", "wikiwords.Rda", package = "largeVis"))
ggplot(words, aes(x = x, y = y, color = label)) + 
  geom_point(size = 0.001, alpha = 0.5) +
  scale_color_gradientn(colors = colors_divergent_discrete(11)(11)) +
  guides(color = FALSE) +
  scale_x_continuous("", limits = c(-3, 5), breaks = NULL) + 
  scale_y_continuous("", limits = c(-4, 4), breaks = NULL) +
  ggtitle(expression(atop("Wiki Word Vectors", atop(paste(sep = " ", "K = 100,", gamma, 
                           "= 7, M = 5,", alpha, "= 1, n_trees = 50,",
                           "max_iters = 1, threshold = 100")))))
```


## Overview of Functions and Hyperparameters

### `randomProjectionTreeSearch`

This function uses a two-phase algorithm to find approximate nearest neighbors. In the first phase, which is based on [Erik Bernhardsson](http://erikbern.com)'s [Annoy](https://github.com/spotify/annoy) algorithm, `n_trees` trees are formed by recursively dividing the space by hyperplanes until at most `tree_threshold` nodes remain in a branch.  A node's candidate nearest neighbors are the union of all nodes with which it shared a leaf on any of the trees.  The `largeVis` algorithm adds a second phase, neighborhood exploration, which considers, for each node, whether the candidate neighbors of the node's candidate immediate neighbors are closer. The logic of the algorithm is that a node's neighbors' neighbors are likely to be the node's own neighbors. In each iteration, the closest `K` candidate neighbors for each node are kept. 

(Note that this implementation of `largeVis` differs from the approach taken by `Annoy`, in that `Annoy` always uses the number of features as the leaf threshold, where `largeVis` allows this to be an adjustable parameter.)

The authors of @tang2016visualizing suggest that a single iteration of the second phase is generally sufficient to obtain satisfactory performance. 

See the vignette "ANN Benchmarks" for additional information.

### `projectKNNs`

This function takes as its input a `Matrix::sparseMatrix`, of connections between nodes. The matrix must be symmetric. A non-zero cell implies that node `i` is a nearest neighbor of node `j`, vice-versa, or both. Non-zero values represent the strength of the connection relative to other nearest neighbors of the two nodes. 

The `LargeVis` algorithm, explained in detail in @tang2016visualizing, estimates the embedding by sampling from the identitied nearest-neighbor connections. For each edge, the algorithm also samples `M` non-nearest neighbor negative samples. `M`, along with $\gamma$ and $\alpha$, control the visualization. $\alpha$ controls the desired distance between nearest neighbors. $\gamma$ controls the relative strength of the attractive force between nearest neighbors and repulsive force between non-neighbors.

The following grid illustrates the effect of the $\alpha$ and $\gamma$ hyperparameters, using the `wiki` dataset which is included with the package:

```{r wikihyperparameters,echo=F,eval=rebuild}
data(wiki)

inputs <- data.frame(
  g = rep(c(.5,1,7,14), 4),
  a = rep(c(.1,1,5,10), each = 4)
)
set.seed(1974)
initialcoords <- matrix(rnorm(ncol(wiki) * 2), nrow = 2)

agcoords <- do.call(rbind, 
                    lapply(1:nrow(inputs), 
                           FUN = function(x) {
  a <- inputs[x, 'a']
  g <- inputs[x, 'g']
  newcoords <- initialcoords
  localcoords <- projectKNNs(wiki, 
                             alpha =  a, 
                             gamma = g,
                             verbose = FALSE, 
                             coords = newcoords)
  localcoords <- data.frame(scale(t(localcoords)))
  colnames(localcoords) <- c("x", "y")
  localcoords$a <- a
  localcoords$g <- g
  localcoords$activity <- log(Matrix::colSums(wiki))
  localcoords  
}))
```
```{r drawhyperparameters,echo=F,fig.width=3.5,fig.height=4,fig.align='center'}
load(system.file("extdata", "agcoords.Rda", package="largeVis"))
ggplot(agcoords,
       aes(x = x, 
           y = y, 
           color = activity)) +
  geom_point(alpha = 0.2, 
             size = 0.05) +
  facet_grid(a ~ g,
             labeller = label_bquote(alpha == .(a), 
                                     gamma == .(g)),
             scales = 'free') +
  scale_x_continuous(breaks = NULL, 
                     name = "") +
  scale_y_continuous(breaks = NULL, 
                     name = "") +
  scale_color_gradientn(colors = colors_continuous(10), 
                        guide=FALSE) +
  ggtitle(expression(paste("Effect of ", alpha, " vs. ", gamma, sep = "  ")))
```

The additional hyperparameters $\rho$ and `min-`$\rho$ control the starting and final learning rate for the stochastic gradient descent process. 

The algorithm can treat positive edge weights in two different ways. The authors of @tang2016visualizing suggest that edge weights should be used to generate a weighted sampling.  However, the algorithm for taking a weighted sample runs in $O(n \log n)$.  Alternatively, the edge-weights can be applied to the gradients.  This is controlled by the `weight_pos_samples` parameter. 

### `vis`

The `vis` function combines `randomProjectionTreeSearch` and `projectKNNs`, along with additional logic for calculating edge weights, to implement the complete `LargeVis` algorithm. 

The following chart illustrates the effect of the `M` and `K` parameters, using the `iris` dataset. Each row re-uses the same set of identified `K` neighbors, and initial coordinates. 

```{r iris,echo=F,fig.width=5,fig.height=5,eval=rebuild}
data(iris)
Ks <- c(5, 10, 20, 30)
Ms <- c(1, 5, 10, 20)
data(iris)
dat <- iris[,1:4]
dupes <- duplicated(dat)
dat <- dat[-dupes,]
labels <- iris$Species[-dupes]
dat <- scale(dat)
dat <- as.matrix(dat)
dat <- t(dat)

set.seed(1974)
coordsinput <- matrix(rnorm(ncol(dat) * 2), nrow = 2)

iriscoords <- do.call(rbind, lapply(Ks, FUN = function(K) {
  neighbors <- randomProjectionTreeSearch(dat, 
                                          K = K, 
                                          verbose = FALSE)
  neighborIndices <- neighborsToVectors(neighbors)
  distances <- largeVis::distance(x = dat, 
                                  neighborIndices$i, 
                                  neighborIndices$j)
  wij <- buildEdgeMatrix(i = neighborIndices$i, 
                       j = neighborIndices$j, 
                       d = distances)
  do.call(rbind, lapply(Ms, FUN = function(M) {
    coords <- projectKNNs(wij = wij$wij, M = M, coords = coordsinput)
    coords <- scale(t(coords))
    coords <- data.frame(coords)
    colnames(coords) <- c("x", "y")
    coords$K <- K
    coords$M <- M
    coords$Species <- as.integer(labels)
    coords
  }))
}))
  
iriscoords$Species <- factor(iriscoords$Species)
levels(iriscoords$Species) <- levels(iris$Species)
```
```{r drawiriscoords,echo=F,fig.width=4,fig.height=4.5,fig.align='center'}
load(system.file("extdata", "iriscoords.Rda", package="largeVis"))
ggplot(iriscoords,
       aes(x = x,
           y = y,
           color = Species)) +
         geom_point(size = 0.5) +
  scale_x_continuous("", 
                     breaks = NULL) +
  scale_y_continuous("", 
                     breaks = NULL) +
  facet_grid(K ~ M, 
             scales = 'free', 
             labeller = label_bquote(K == .(K), M == .(M))) +
  scale_color_manual(values = colors_discrete(3)) +
  ggtitle("Effect of M and K on Iris Dataset")
```

### `manifoldMap`

The `manifoldMap` function is useful when the examples being clustered are themselves images. Given a coordinate matrix (as generated by `projectKNNs` or `vis`) and an `array` of `N` images, the function samples `n` images and plots them at the coordinates given in the matrix. 

The function can plot both color and greyscale images. 

#### Example with MNIST Letters

The following code will plot 5000 images sampled from the MNIST dataset at positions generated by `vis`:
```{r loadmnistimages,eval=rebuild,echo=F}
load("data/train.RData")
```
```{r drawmanifoldmap,echo=T,fig.width=8,fig.height=8,message=F,warning=F,fig.align='center'}
par(mai=c(0.25, 0.25, 0.25, 0.25))
if (exists("trainData")) {
  dim(trainData) <- c(60000, 28, 28)
  set.seed(1974)
  manifoldMap(mnistCoords[,1:2],
      n = 5000,
      scale = 0.003,
      images = trainData,
      xlab = "", 
      ylab = "",
      xlim = c(-2, 2),
      ylim = c(-2, 2))
} 
```
```{r nomnist,eval=!buildManifolds,echo=F,results='asis'}
cat("The plot is disabled by default because it requires the MNIST dataset.  To recreate the plot, change the vignette script to point to the downloaded images.\n")
cat("The MNIST data may be obtained using the `darch` package, available on CRAN, with the commands `provideMNIST(folder = 'download location', download = TRUE)` followed by `readMNIST(folder = 'download location')`")
```

#### Example with Faces

The following examples visualize facial-recognition embedding vectors from the [Labelled Faces in the Wild](http://vis-www.cs.umass.edu/lfw/) dataset.  The embedding vectors were graciously provided by [Brandon Amos](https://bamos.github.io/) of the [OpenFace](https://cmusatyalab.github.io/openface/) project. Similar vectors may be generated for images using the OpenFace `batch-represent` command.

```{r lfw,echo=F,eval=rebuild}
data("facevectors")

faceembeddings <- t(as.matrix(facevectors[,-c(1:2)]))
faceVis <- vis(faceembeddings, K = 50, 
               max_iter = 2, 
               n_trees = 100, 
               distance_method = 'Cosine')

faceCoords <- data.frame(scale(t(faceVis$coords)))
colnames(faceCoords) <- c("x", "y")
faceCoords <- cbind(faceCoords, facevectors[,1:2])
```

OpenFace embedding vectors encode an image in such a way that the embeddings for multiple images of the same person should be similar.  This is illustrated on the first plot below, which highlights the locations of the embedding vectors for images of 10 selected individuals. 

```{r plotFaceVectors,echo=F,fig.width=7,fig.height=5}
load(system.file("extdata", "faces.Rda", package="largeVis"))
faceCoCopy <- faceCoords
lvs <- c("Tony_Bennette", 
         "Gloria_Gaynor", 
         "Jennifer_Aniston", 
         "Kobe_Bryant", 
         "John_Nash",
         "Jack_Smith", 
         "Nancy_Kerrigan", 
         "Nora_Ephron",
         "Julianna_Margulies", 
         "Abdullah_al-Attiyah")
faceCoCopy$name[! faceCoCopy$name %in% lvs] <- "Other"
faceCoCopy$name <- factor(faceCoCopy$name)
faceCoCopy$alpha <- factor(ifelse(faceCoCopy$name == "Other", 0.05, 0.2))

ggplot(faceCoCopy, aes(x = x, 
                       y = y, 
                       color = name,
                       alpha = alpha,
                       size = alpha)) + 
  geom_point() +
  scale_alpha_manual(values = c(0.2, 0.8), guide = FALSE) + 
  scale_color_manual(values = c("grey", colors_divergent_discrete(10)(10))) +
  scale_size_manual(values = c(0.2, 0.5), guide = FALSE) +
  scale_x_continuous("", 
                     breaks = NULL) +
  scale_y_continuous("", 
                     breaks = NULL) +
  ggtitle("Positions of Embedding Vectors for Selected Individuals")
```

The function of `manifoldMap` is illustrated in the following plot, which places 500 images from the dataset at the locations given by the `largeVis` map.  

```{r faceImages,eval=buildManifolds,echo=F,fig.width=8,fig.height=8}
load(system.file("extdata", "faceLabels.Rda", package="largeVis"))
set.seed(1974)
n <- 500
facesToPlot <- sample(nrow(faceCoords), n)

faces <- apply(faceLabels[facesToPlot,], MARGIN = 1, FUN = function(x) {
  ret <- readJPEG(paste("/mnt/hfsshare/DATASETS/lfw faces/lfw",
                x[1], sub("png", "jpg", x[2]), sep = "/"))
  dim(ret) <- c(250, 250, 3)
  ret
})
dim(faces) <- c(250, 250, 3, n)
faces <- aperm(faces, c(4,1,2,3)) # n,h,w,c
ggManifoldMap(
  x = faceCoords[facesToPlot,1:2], 
  n = n, 
  images = 1 - faces, 
  scale =  1 / 1000) +
  scale_y_continuous(name = "", breaks = NULL) +
  scale_x_continuous(name = "", breaks = NULL) + 
  ggtitle("Manifold Map of OpenFace Embeddings")
```

```{r faceImages5000,eval=buildManifolds,echo=F,results='asis'}
png(filename = "faceshighres.png", 
    width = 5000, height = 5000, units = 'px', 
    bg = 'grey5')
manifoldMap(x = faceCoords[facesToPlot,1:2], 
            n = n, images = 1 - faces, scale =  1 / 1000,
            xlab = NULL, ylab = NULL, col.lab = 'gray5',
            col.axis = 'gray5')
cat("A high resolution version is available [here](vignettes/faceshighres.png)")
```

```{r noface,eval=!buildManifolds,results='asis',echo=F}
cat("The plot is disabled by default because it requires face images from [Labeled Faces in the Wild](http://vis-www.cs.umass.edu/lfw/). To recreate the plot, change the vignette script to point to the downloaded images.")
```
Note what the plot is showing. The positions of the faces in [Karpathy's embedding plots](http://cs.stanford.edu/people/karpathy/cnnembed/) are highly sensitive to the background and overall sense of each image, because Karpathy used embedding vectors produced by a general-purpose model. In the plot of the OpenFace vectors above, the positions are insensitive to the image backgrounds -- but pictures of the same individual are plotted close to each other, and close to similar-looking persons. 

## Support for Sparse Matrices

`largeVis` supports sparse matrices.  Besides facilitating very large datasets, this makes it practicable to visualize term-document-matrices. 

For example, the following plot visualizes a corpus of 5000 political blog entries, as included with the `stm` package.  The first row visualizes the blog entries as tf-idf weighted term vectors; the second, as topic vectors.

```{r stm,echo=F,eval=rebuild}
library(stm)
data("poliblog5k")
p <- c(0, cumsum(as.numeric(lapply(poliblog5k.docs, function(x) ncol(x)))))
i <- do.call("c", lapply(poliblog5k.docs, function(x) x[1,]))
p[length(p)] <- length(i)
j <- rep(0:(length(diff(p)) - 1), diff(p))
v <- do.call("c", lapply(poliblog5k.docs, function(x) x[2,]))
poli <- Matrix::sparseMatrix(i = i + 1, 
                             j = j + 1, 
                             x = v)
dupes <- duplicated(slam::as.simple_triplet_matrix(Matrix::t(poli)))
poli <- poli[, ! dupes]
poli <- poli / log(Matrix::rowSums(poli > 0)) # tf-idf weight
set.seed(1974)
inputcoords <- matrix(rnorm(10000), nrow = 2)
policoords <- vis(poli, 
                  K = 100, 
                  n_trees = 50, 
                  tree_threshold = 100, 
                  max_iter = 1,
                  M = 10,
                  distance_method = 'Cosine',
                  coords = inputcoords,
                  verbose = FALSE)

stmmodel <- stm(poliblog5k.docs, poliblog5k.voc, K = 20, 
                data = poliblog5k.meta, prevalence = ~ rating + s(day),
                content = ~rating,
                max.em.its = 75, init.type="Spectral", seed = 1974)

stmvectors <- t(scale(stmmodel$theta))

set.seed(1974)
inputcoords <- matrix(rnorm(10000), nrow = 2)
stmVis <- vis(stmvectors, 
                  K = 100, 
                  n_trees = 50, 
                  tree_threshold = 100, 
                  max_iter = 1,
                  M = 10,
                  distance_method = 'Cosine',
                  coords = inputcoords,
                  verbose = FALSE)

polidata <- data.frame(scale(t(policoords$coords)))
colnames(polidata) <- c('x', 'y')
polidata$rating <- poliblog5k.meta$rating[!dupes]
polidata$blog <- poliblog5k.meta$blog[!dupes]

stmdata <- data.frame(scale(t(stmVis$coords)))
colnames(stmdata) <- c('x', 'y')
stmdata$rating <- poliblog5k.meta$rating
stmdata$blog <- poliblog5k.meta$blog

polidata$origin <- "tf-idf term vectors"
stmdata$origin <- "stm topic vectors"

combined <- rbind(polidata, 
                  stmdata)
combined$origin <- factor(combined$origin)
combined$origin <- factor(combined$origin, 
                          levels = rev(levels(combined$origin)))
```


```{r drawtdm,echo=F,fig.height=4,fig.width=7,warning=FALSE,message=FALSE,error=FALSE}
load(system.file("extdata", "poliblog.Rda", package="largeVis"))

ggplot(combined, aes(x = x, 
                     y = y, 
                     color = blog)) +
  geom_point(size = 0.2, 
             alpha = 0.8) +
  scale_color_manual(values = colors_divergent_discrete(6)(6)) +
  facet_grid(origin ~ rating, 
             scale = 'free') +
  scale_x_continuous("", 
                     breaks = NULL, 
                     limits = c(-2.5, 2.5)) +
  scale_y_continuous("", 
                     breaks = NULL,
                     limits = c(-2.5, 2.5)) +
  ggtitle("5000 Political Blog Entries")
```

This facilitates evaluation of the effectiveness of a topic model.  In the example above, `stm` is asked to create a 20-topic model that is aware that blogs marked "liberal" and blogs marked "conservative" may talk about the same topics in different ways.  As expected, the `largeVis` visualization of this model shows blog entries that group into 20 clusters, with "liberal" and "conservative" blogs each addressing, but placing different weights on, each of the 20 topics.

## Visualizing Graphs

The `largeVis` visualization algorithm can be used to visualize ordinary graphs.  The included `wiki` dataset is an example.

The following code illustrates how to import and visualize a graph using the YouTube-communities dataset available [here](https://snap.stanford.edu/data/com-Youtube.html). The data and visualization are not included here for size reasons.

```{r youtube,eval=F,echo=T}
pathToGraphFile <- 
  "/mnt/hfsshare/DATASETS/YouTubeCommunities/com-youtube.ungraph.txt"
pathToCommunities <- 
  "/mnt/hfsshare/DATASETS/YouTubeCommunities/com-youtube.top5000.cmty.txt"

youtube <- readr::read_tsv(pathToGraphFile, skip=4, col_names=FALSE)
youtube <- as.matrix(youtube)
youtube <- Matrix::sparseMatrix(i = youtube[, 1],
                                j = youtube[, 2],
                                x = rep(1, nrow(youtube)), 
                                dims = c(max(youtube), max(youtube)))
youtube <- youtube + t(youtube)
communities <- readr::read_lines(pathToCommunities)
communities <- lapply(communities, 
                      FUN = function(x) as.numeric(unlist(strsplit(x, "\t"))))
community_assignments <- rep(0, 
                             nrow(youtube))
for (i in 1:length(communities)) community_assignments[communities[[i]]] <- i

youTube_coordinates <- projectKNNs(youtube)
youTube_coordinates <- data.frame(scale(t(youTube_coordinates)))
colnames(youTube_coordinates) <- c("x", "y")
youTube_coordinates$community <- factor(community_assignments)
youTube_coordinates$alpha <- factor(ifelse(youTube_coordinates$community == 0, 0.05, 0.2))
ggplot(youTube_coordinates, aes( x = x, 
                      y = y, 
                      color = community, 
                      alpha = alpha, 
                      size = alpha)) +
  geom_point() +
  scale_color_manual(values = 
                       c("black", colors_continuous(5000)),
                     guide = FALSE) +
  scale_alpha_manual(values = c(0.005, 0.2), guide = FALSE) +
  scale_size_manual(values = c(0.03, 0.15), guide = FALSE) +
  scale_x_continuous("", 
                     breaks = NULL, limits = c(-2.5,2.5)) +
  scale_y_continuous("", 
                     breaks = NULL, limits = c(-2.5,2.5)) +
  ggtitle("YouTube Communities")
```

## Distance Methods

The original `LargeVis` paper used Euclidean distances exclusively.  The `largeVis` package offers a choice among Euclidean and Cosine distance measures.  

The implementation is not optimized for cosine distances. Using cosine distances will definitely be
slower than using Euclidean distances. This is because the random projection tree algorithm for cosine distances requires that the input vectors in the data matrix be normalised. 

## Memory Consumption

The algorithm is necessarily memory-intensive for large datasets. 

A simple way to reduce peak memory usage, is to turn-off the `save_neighbors` and `save_sigmas` parameters when running `vis`. 

If this is insufficient, the steps of the algorithm can be run separately with the `neighborsToVectors`, `distance`, and `buildEdgeMatrix` functions.  In this case, the workflow is:

```{r lowmemexample,eval=F,echo=T}
neighbors <- randomProjectionTreeSearch(largeDataset)
neighborIndices <- neighborsToVectors(neighbors)
rm(neighbors)
gc()
distances <- distance(x = largeDataset, 
                      i = neighborIndices$i, 
                      j =neighborIndices$j)
rm(largeDataset)
gc()
wij <- buildEdgeMatrix(i = neighborIndices$i, 
                       j = neighborIndices$j, 
                       d = distances)
rm(distances, neighborIndices)
gc()
coords <- projectKNNs(wij$wij)
```

Note that `gc()` is being called explicitly. The reason is that R will not garbage collect while executing the package's C++ functions, which can require substantial temporary RAM. 

In testing, this method reduced peak RAM requirements by more than 70%. 

Memory requirements during the neighbor search may be managed by reducing `n_trees` and increasing the `tree_threshold`. The decrease in precision is marginal, and may be compensated-for by increasing `max_iters`.  See the benchmarks vignette for further detail.

## References
